---
title: "Performance Evaluation of Early-Stage Dementia Prediction Models"
author: "Linglu Li"
output:
  pdf_document:
    latex_engine: xelatex  
    toc: false
    number_sections: false
    fig_width: 6
    fig_height: 3
---

```{r, include=FALSE}
library(tidyr)
library(readxl)
library(MASS)
library(dplyr) 
library(purrr)
library(ggplot2)
library(tibble)
library(knitr)
library(kableExtra)
library(patchwork)
library(VIM)
library(tree)
library(reshape2)
library(randomForest)
library(gbm)
library(ISLR)
library(caret)
library(xgboost)
library(data.table)
```
The target variable is well-balanced across both cohorts, with the cross-sectional dataset (cross2_imp) showing a 57:43 distribution and the longitudinal dataset (df_2group) achieving a nearly equal 48:52 split, thereby mitigating the risk of majority-class bias and the Accuracy Paradox.
```{r, include = FALSE}

longitutional <- read.csv("/Users/tera/Desktop/PSTAT/131/project/longitudinal.csv")

longi_one_row <- longitutional %>%
    group_by(Subject.ID) %>%
    slice_max(order_by = Visit, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(Group2 = ifelse(Group %in% c("Converted", "Demented"), 
                           "Demented", "Nondemented"),
           Group2 = factor(Group2, levels = c("Nondemented", "Demented")))
set.seed(1)
train_idx_long <- sample(nrow(longi_one_row), 0.75 * nrow(longi_one_row))
train_long_raw <- longi_one_row[train_idx_long, ]
test_long_raw  <- longi_one_row[-train_idx_long, ]
ses_median_long  <- median(train_long_raw$SES, na.rm = TRUE)
mmse_median_long <- median(train_long_raw$MMSE, na.rm = TRUE)

prepare_long_data <- function(data, ses_val, mmse_val) {
  data %>%
    # Fill missing values with the Training anchor
    mutate(
      SES  = ifelse(is.na(SES),  ses_val,  SES),
      MMSE = ifelse(is.na(MMSE), mmse_val, MMSE)
    ) %>%
    mutate(Educ = case_when(
      EDUC <= 11      ~ 1,
      EDUC == 12      ~ 2,
      EDUC %in% 13:15 ~ 3,
      EDUC == 16      ~ 4,
      EDUC >= 17      ~ 5
    )) %>%
    mutate(Educ = factor(Educ, ordered = TRUE)) %>%
    # Drop unused clinical/ID columns
    select(-EDUC, -Subject.ID, -MRI.ID, -Group, -Visit, -MR.Delay, -Hand, -CDR)
}

train_long <- prepare_long_data(train_long_raw, ses_median_long, mmse_median_long)
test_long  <- prepare_long_data(test_long_raw, ses_median_long, mmse_median_long)
```

```{r, include = FALSE}
cross <- read.csv("/Users/tera/Desktop/PSTAT/131/project/oasis_cross-sectional.csv")
cross2 <- cross %>%
  mutate(
    Group2 = ifelse(CDR == 0, "Nondemented", "Demented"),
    Group2 = factor(Group2, levels = c("Nondemented", "Demented"))
  ) %>%
  select(M.F, Age, Educ, SES, MMSE, eTIV, nWBV, ASF, Group2)%>%
  filter(!is.na(Group2))

set.seed(1)

train_idx <- sample(nrow(cross2), 0.75 * nrow(cross2))
train_raw <- cross2[train_idx, ]
test_raw  <- cross2[-train_idx, ]

ses_median_anchor <- median(train_raw$SES, na.rm = TRUE)

train_dat <- train_raw %>%
  mutate(SES = ifelse(is.na(SES), ses_median_anchor, SES))

test_dat <- test_raw %>%
  mutate(SES = ifelse(is.na(SES), ses_median_anchor, SES))


```


```{r splitting, include = FALSE}

y_longtrain = train_long$Group2
y_longtest = test_long$Group2

train_dat$Group2 <- relevel(train_dat$Group2, ref = "Demented")
test_dat$Group2  <- relevel(test_dat$Group2,  ref = "Demented")
y_train <- train_dat$Group2
y_test <- test_dat$Group2

set.seed(1)
folds_cross <- createFolds(train_dat$Group2, k = 10)
folds_long <- createFolds(train_long$Group2, k = 10)

# Set num of folds
train_control_cross <- trainControl(
  method = "cv",
  number = 10,
  verboseIter = TRUE,
  classProbs = TRUE,
  summaryFunction = twoClassSummary, # Adds AUC/Sensitivity/Specificity
  savePredictions = "final",
  index = folds_cross
)

train_control_long <- trainControl(
  method = "cv",
  number = 10,
  repeats = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final",
  index = folds_long
)
```

```{r, include=FALSE}

set.seed(1)
logit_model <- train(
  Group2 ~ .,
  data      = train_dat,
  method    = "glmnet",
  family    = "binomial",
  trControl = train_control_cross,
  metric  = "ROC" 
)


logit_model1 <- train(
  Group2 ~ .- eTIV - nWBV - ASF,
  data      = train_dat,
  method    = "glmnet",
  family    = "binomial",
  trControl = train_control_cross,
  metric  = "ROC"
)


logit_model2 <- train(
  Group2 ~ .,
  data      = train_long,
  method    = "glmnet",
  family    = "binomial",
  trControl = train_control_long,
  metric  = "ROC")

logit_model3 <- train(
  Group2 ~ .- eTIV - nWBV - ASF,
  data      = train_long,
  method    = "glmnet",
  family    = "binomial",
  trControl = train_control_long,
  metric  = "ROC"
)

```

```{r, include=FALSE}
logit_pred <- predict(logit_model, test_dat)
logit_conf <- confusionMatrix(logit_pred, test_dat$Group2)
logit_cv_acc <- max(logit_model$results$Accuracy)
logit_pred1 <- predict(logit_model1, test_dat)
logit_pred2 <- predict(logit_model2, test_long)
logit_pred3 <- predict(logit_model3, test_long)
```


```{r include = FALSE }
# with two group notused


#draw.tree(tree.long2, nodeinfo=TRUE, cex = 0.4)
# title("Classification Tree Predicting Dementia Status (2 groups)", cex = 0.8)

tree.longtrain2 <- tree(Group2~., data = train_long)
  
# Cross-Validation for tree
cv_tree2 <- cv.tree(tree.longtrain2, FUN = prune.misclass, K = 5)
plot(cv_tree2)

best_size2 <- cv_tree2$size[which.min(cv_tree2$dev)]


# Plot size vs. cross-validation error rate
plot(cv_tree2$size , cv_tree2$dev, type="b", 
     xlab = "Number of leaves, \'best\'", ylab = "CV Misclassification Error",
     col = "red", main="CV (2-group)")
abline(v=best_size2, lty=2)
# Add lines to identify complexity parameter
min.error2 = which.min(cv_tree2$dev) # Get minimum error index
abline(h = cv_tree2$dev[min.error2],lty = 2)

pruned_tree2 <- prune.misclass(tree.longtrain2, best=best_size2)
plot(pruned_tree2)
text(pruned_tree2, pretty = 0, col = "blue", cex = .5)
title("Pruned tree of size 2")

```

```{r DTmodel,include = FALSE}
tree_grid <- expand.grid(cp = seq(0, 0.02, length.out = 50))

set.seed(1)
tree_model <- train(
  Group2 ~ .,
  data = train_dat,
  method = "rpart",
  trControl = train_control_cross,
  tuneGrid  = tree_grid, 
  metric = "ROC")

tree_model1 <- train(
  Group2 ~ . - eTIV - nWBV - ASF,
  data = train_dat,
  method = "rpart",
  trControl = train_control_cross,
  tuneGrid  = tree_grid,
  metric = "ROC"
  )

tree_model2 <- train(
  Group2 ~ . ,
  data = train_long,
  method = "rpart",
  trControl = train_control_long,
  tuneGrid  = tree_grid,
  metric = "ROC"
  )

tree_model3 <- train(
  Group2 ~ . - eTIV - nWBV - ASF,
  data = train_long,
  method = "rpart",
  trControl = train_control_long,
  tuneGrid  = tree_grid, 
  metric = "ROC")

```

```{r tree-conf, include = F}
tree_pred <- predict(tree_model, newdata = test_dat)
tree_conf <- confusionMatrix(tree_pred, test_dat$Group2)
tree_acc <- tree_conf$overall["Accuracy"]
best_cp <- tree_model$bestTune$cp
tree_cv_acc <- tree_model$results$Accuracy[
  tree_model$results$cp == best_cp]

tree_pred1 <- predict(tree_model1, newdata = test_dat)
tree_pred2 <- predict(tree_model2, newdata = test_long)
tree_pred3 <- predict(tree_model3, newdata = test_long)
```



```{r rf-fold,include= FALSE}
set.seed(1)

rf_model <- train(
  Group2 ~ . ,
  data      = train_dat,
  method    = "rf",
  trControl = train_control_cross,
  tuneGrid   = data.frame(mtry = c(1, 2, 3, 4, 5)), 
  importance = TRUE,
  metric = "ROC")

rf_model1 <- train(
  Group2 ~ . - eTIV - nWBV - ASF,
  data      = train_dat,
  method    = "rf",
  trControl = train_control_cross,
  tuneGrid   = data.frame(mtry = c(1, 2, 3, 4, 5)), 
  importance = TRUE,
  metric = "ROC")

rf_model2 <- train(
  Group2 ~ .,
  data      = train_long,
  method    = "rf",
  trControl = train_control_long,
  tuneGrid   = data.frame(mtry = c(1, 2, 3, 4, 5)), 
  importance = TRUE,
  metric = "ROC")

rf_model3 <- train(
  Group2 ~ . - eTIV - nWBV - ASF,
  data      = train_long,
  method    = "rf",
  trControl = train_control_long,
  tuneGrid   = data.frame(mtry = c(1, 2, 3, 4, 5)), 
  importance = TRUE,
  metric = "ROC")
```

```{r rf-conf, include = F}
rf_pred <- predict(rf_model, newdata = test_dat)
rf_cv_acc <- rf_model$results$Accuracy[ rf_model$results$mtry == rf_model$bestTune$mtry ] # CV accuracy

rf_conf <- confusionMatrix(rf_pred, test_dat$Group2)
rf_acc <- rf_conf$overall["Accuracy"]

rf_pred1 <- predict(rf_model1, newdata = test_dat)
rf_pred2 <- predict(rf_model2, newdata = test_long)
rf_pred3 <- predict(rf_model3, newdata = test_long)
```


```{r RandomForest,  include = F}
# NotUsed
set.seed(1)

rf_model0 <- randomForest(
  Group2 ~ .,
  data      = train_long,
  ntree     = 500,    # number of trees
  mtry      = 3,      # number of variables tried at each split (you can tune this)
  importance = TRUE   # to look at variable importance later
)

# varImpPlot(rf_model)
```


```{r XGboosting, include = FALSE}

grid_tune <- expand.grid(
  nrounds = c(50, 100, 150), # number of trees
  max_depth = c(2,3,4),
  eta = c(0.01, 0.1, 0.3), #learning rate
  gamma = c(0,1), # pruning draft
  colsample_bytree = c(0.6, 0.8), # % of features used per tree
  min_child_weight = c(1), # the larger the more conservative the model; can be used as a stop
  subsample = 0.75 # used to prevent overfitting by sampling x% train
)

# Drop factor predictors that have only 1 level
# is_bad_factor <- sapply(train_dat, function(v) is.factor(v) && nlevels(v) < 2)
# train_dat <- train_dat[, !is_bad_factor]
```

```{r xgb-fold,include = FALSE}
set.seed(1)
xgb_model <- train(
  Group2 ~ . ,
  data      = train_dat,
  method    = "xgbTree",
  trControl = train_control_cross,
  tuneGrid  = grid_tune,
  verbose   = TRUE, 
  metric = "Accuracy"
)

xgb_model1 <- train(
  Group2 ~ . - eTIV - nWBV - ASF,
  data      = train_dat,
  method    = "xgbTree",
  trControl = train_control_cross,
  tuneGrid  = grid_tune,
  verbose   = TRUE,
  metric = "Accuracy"
)

xgb_model2 <- train(
  Group2 ~ . ,
  data      = train_long,
  method    = "xgbTree",
  trControl = train_control_long,
  tuneGrid  = grid_tune,
  verbose   = TRUE,
  metric = "Accuracy"
)

xgb_model3 <- train(
  Group2 ~ . - eTIV - nWBV - ASF,
  data      = train_long,
  method    = "xgbTree",
  trControl = train_control_long,
  tuneGrid  = grid_tune,
  verbose   = TRUE,
  metric = "Accuracy"
)
```

```{r xgb-conf, include = FALSE}
xgb_pred <- predict(xgb_model, newdata = test_dat)
xgb_conf <- confusionMatrix(xgb_pred, test_dat$Group2)
xgb_accu <- xgb_conf$overall["Accuracy"]

best_pars <- xgb_model$bestTune
xgb_cv_acc <- xgb_model$results$Accuracy[
  apply(xgb_model$results[, names(best_pars)], 1, function(row)
    all(row == best_pars))
]
xgb_pred1 <- predict(xgb_model1, newdata = test_dat)
xgb_pred2 <- predict(xgb_model2, newdata = test_long)
xgb_pred3 <- predict(xgb_model3, newdata = test_long)


```





```{r SVM, include = FALSE}
set.seed(1)

svm_grid <- expand.grid(
  C     = c(0.25,0.5,1),
  sigma = c(0.005, 0.01, 0.02)
)

svm_control <- trainControl(
  method = "repeatedcv",
  number = 15,
  repeats = 10,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)



svm_model <- train(
  Group2 ~ . ,
  data       = train_dat,
  method     = "svmRadial",
  trControl  = train_control_cross,
  preProcess = c("center", "scale"), 
  tuneGrid   = svm_grid,
  metric     = "ROC"
  
)


svm_model1 <- train(
  Group2 ~ . - eTIV - nWBV - ASF,
  data       = train_dat,
  method     = "svmRadial",
  trControl  = train_control_cross,
  preProcess = c("center", "scale"), 
  tuneGrid   = svm_grid,
  metric     = "ROC"
)


svm_model2 <- train(
  Group2 ~ . ,
  data       = train_long,
  method     = "svmRadial",
  trControl  = train_control_long,
  preProcess = c("center", "scale", "pca"), 
  tuneGrid   = svm_grid,
  metric     = "ROC"
)


svm_model3 <- train(
  Group2 ~ . - eTIV - nWBV - ASF,
  data       = train_long,
  method     = "svmRadial",
  trControl  = train_control_long,
  preProcess = c("center", "scale"), 
  tuneGrid   = svm_grid,
  metric     = "ROC"
)
```

```{r, include = FALSE}
svm_pred <- predict(svm_model, newdata = test_dat)
svm_conf <- confusionMatrix(svm_pred, test_dat$Group2)
svm_acc <- svm_conf$overall["Accuracy"]
best_C     <- svm_model$bestTune$C
best_sigma <- svm_model$bestTune$sigma

svm_cv_acc <- svm_model$results$Accuracy[
  svm_model$results$C == best_C &
  svm_model$results$sigma == best_sigma
]

svm_pred1 <- predict(svm_model1, newdata = test_dat)
svm_pred2 <- predict(svm_model2, newdata = test_long)
svm_pred3 <- predict(svm_model3, newdata = test_long)
```



# Evaluation Metrics

The Table 1 shows the performance of five different classifier models: Decision Tree, Random Forest, Support Vector Machine (SVM), XGBoost, and Logistic Regression. The comparison is based on the following metrics:

- **Accuracy**: The proportion of correctly classified result from the total instances.
\
- **Precision**: The ratio of correctly predicted positive rate to the total predicted positive rates. A good classifier close to value 1.
\
- **Recall**: The true positive rate. A good classifier close to value 1.
\
* **F1 Score**: The weighted average of Recall and Precision, which balances both metrics.

The SVM and Logistic Regression models performed the worst in identifying actual dementia patients, as indicated by their lowest Recall of $0.75$. This means they only correctly identified $75\%$ of the actual dementia cases in the test set, leading to the highest proportion of misclassified dementia patients. The model best at correctly identifying actual dementia cases is XGBoost with a Recall of $0.88$. 

XGBoost also ties with the Decision Tree for the highest overall Accuracy ($89.36\%$), with each model reflecting a beneficial aspect of the Precision–Recall trade-off. The Decision Tree offers higher Precision ($0.87$), effectively reducing False Positives (i.e., fewer healthy patients are misdiagnosed). XGBoost, on the other hand, excels at capturing true dementia cases, reaching a Recall of ($0.88$) while maintaining solid Precision ($0.82$).

XGBoost is the recommended model because its high Recall minimizes the risk of missed diagnoses, making it well-suited for clinical decision-making. Although its Precision is slightly lower, its strong ability to avoid false negatives makes it the safest and most reliable choice for detecting disease.

```{r, echo=FALSE, message=FALSE, warning =FALSE}
compute_metrics <- function(pred, truth) {
  
  cm <- confusionMatrix(pred, truth, positive = "Demented")
  
  acc  <- cm$overall["Accuracy"]
  prec <- cm$byClass["Precision"]
  rec  <- cm$byClass["Recall"]
  f1   <- cm$byClass["F1"]
  
  return(c(Accuracy = acc,
           Precision = prec,
           Recall = rec,
           F1 = f1))
}

# Use the actual test labels from test_dat
tree_metrics  <- compute_metrics(tree_pred,  test_dat$Group2)
rf_metrics    <- compute_metrics(rf_pred,    test_dat$Group2)
svm_metrics   <- compute_metrics(svm_pred,   test_dat$Group2)
xgb_metrics   <- compute_metrics(xgb_pred,   test_dat$Group2)
logit_metrics <- compute_metrics(logit_pred, test_dat$Group2)

results_table <- rbind(
  "Decision Tree" = tree_metrics,
  "Random Forest" = rf_metrics,
  "SVM"           = svm_metrics,
  "XGBoost"       = xgb_metrics,
  "Logistic Reg." = logit_metrics
)

colnames(results_table) <- c("Accuracy", "Precision", "Recall", "F1")
results_print <- as.data.frame(results_table)

# Round Precision, Recall, F1-score to 2 decimals
results_print$Accuracy  <- round(results_print$Accuracy, 4)
results_print$Precision <- round(results_print$Precision, 2)
results_print$Recall    <- round(results_print$Recall, 2)
results_print$F1        <- round(results_print$F1, 2)

# Convert Accuracy to percentage notation
results_print[, "Accuracy"] <- sprintf("%.2f%%", results_print[, "Accuracy"] * 100)


```
```{r, include=FALSE}
# Function to calculate, format, and display model results
generate_performance_report <- function(pred_list, truth, title = "Performance comparison of ML models.") {
  
  # 1. Apply your compute_metrics function to every prediction in the list
  # This handles tree_pred, tree_pred1, etc., automatically
  metrics_list <- lapply(pred_list, function(p) {
    # Ensure prediction and truth have same length
    if(length(p) != length(truth)) return(rep(NA, 4)) 
    compute_metrics(p, truth)
  })
  
  # 2. Combine into a matrix and then a dataframe
  results_df <- as.data.frame(do.call(rbind, metrics_list))
  
  # 3. Formatting (Rounding and Percentages)
  # Keeping 4 decimals for Accuracy before percentage conversion
  if(ncol(results_df) == 4) {
    colnames(results_df) <- c("Accuracy", "Precision", "Recall", "F1")
  }
  
  # 4. Final formatting for the Writing Sample
  results_df %>%
    mutate(
      Accuracy = sprintf("%.2f%%", round(Accuracy, 4) * 100),
      across(c(Precision, Recall, F1), ~round(.x, 2))
    ) %>%
    kable(caption = title, align = "c") %>%
    kable_styling(full_width = FALSE, position = "center") %>%
    row_spec(0, bold = TRUE)
}




```

Linear classifiers (logistic regression and SVM) demonstrated higher accuracy in NoMRI models, likely due to reduced multicollinearity and improved linear separability among clinical predictors. In contrast, tree-based methods benefited more from MRI-derived variables, which introduce non-linear and interaction effects that are better captured by hierarchical splits.

```{r, echo=FALSE}

full_model_preds <- list(
  "Decision Tree" = tree_pred,
  "Random Forest" = rf_pred,
  "SVM"           = svm_pred,
  "XGBoost"       = xgb_pred,
  "Logistic Reg." = logit_pred
)

generate_performance_report(full_model_preds, test_dat$Group2, "Performance comparison of Full ML Models - CrossSection") 

#Table without MRI variables
half_model_preds <- list(
  "Decision Tree" = tree_pred1,
  "Random Forest" = rf_pred1,
  "SVM"           = svm_pred1,
  "XGBoost"       = xgb_pred1,
  "Logistic Reg." = logit_pred1
)

generate_performance_report(half_model_preds, test_dat$Group2, "Performance comparison of NoMRI ML Models - CrossSection") 

longf_model_preds <- list(
  "Decision Tree" = tree_pred2,
  "Random Forest" = rf_pred2,
  "SVM"           = svm_pred2,
  "XGBoost"       = xgb_pred2,
  "Logistic Reg." = logit_pred2
)

generate_performance_report(longf_model_preds, test_long$Group2, "Performance comparison of Full ML Models - Longitutional") 

#Table without MRI variables
longh_model_preds <- list(
  "Decision Tree" = tree_pred3,
  "Random Forest" = rf_pred3,
  "SVM"           = svm_pred3,
  "XGBoost"       = xgb_pred3,
  "Logistic Reg." = logit_pred3
)

generate_performance_report(longh_model_preds, test_long$Group2, "Performance comparison of NoMRI ML Models - Longitutional") 
```

\
Note: A positive prediction indicates that the model predicts the patient has Alzheimer’s Disease (is Demented), whereas a negative prediction indicates the model predicts the patient does not have Alzheimer’s Disease (is Nondemented). All four evaluation metrics—Accuracy, Precision, Recall, and F1-score—are calculated from the fundamental confusion matrix components (TP, TN, FP, and FN). More detailed definitions of these terms are provided in Tables 2 and 3.
\newpage 



```{r prediction-tables, echo=FALSE, message=FALSE, warning=FALSE, fig.weight= 16, fig.height = 5, fig.align='center'}


# 1. Data for Correct Predictions Table
correct_predictions <- data.frame(
  Term = c("TP", "TN"),
  FullName = c("True Positive", "True Negative"),
  ClinicalContext = c(
    "The model predicts the patient is Demented, and the patient actually is Demented. (A successful diagnosis).",
    "The model predicts the patient is Nondemented, and the patient actually is Nondemented. (A successful exclusion of the disease)."
  ),
  stringsAsFactors = FALSE
)

# 2. Data for Incorrect Predictions (Errors) Table
incorrect_predictions <- data.frame(
  Term = c("FP", "FN"),
  FullName = c("False Positive", "False Negative"),
  ClinicalContext = c(
    "The model predicts the patient is Demented, but the patient actually is Nondemented. (A misdiagnosis).",
    "The model predicts the patient is Nondemented, but the patient actually is Demented. (A missed diagnosis)."
  ),
  stringsAsFactors = FALSE
)

# 3.  (Correct Predictions)
kable(
  correct_predictions,
  caption = "Correct Outcomes from the Confusion Matrix.",
  align = "c",
  col.names = c("Term", "Full Name", "Clinical Context")
) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  column_spec(3, width = "25em") %>% 
  row_spec(0, bold = TRUE, background = "#D9EAD3")

# 4. Create and style the second table (Incorrect Predictions)

kable(
  incorrect_predictions,
  caption = "Incorrect Outcomes (Errors) from the Confusion Matrix.",
  align = "c",
  col.names = c("Term", "Full Name", "Clinical Context")
) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  column_spec(3, width = "25em") %>% 
  row_spec(0, bold = TRUE, background = "#F9C8C8")
```

# Performance Table Visualization
Detailed comparisons of model performance metrics (Accuracy, Precision, Recall, and F1-score) are visually represented in Figure 1.
\

```{r,echo = FALSE, message=FALSE, warning =FALSE, fig.height= 10, fig.width= 16, fig.cap = "Model comparison across Accuracy, Precision, Recall, and F1-score."}
results <- data.frame(
  Model = c("Decision Tree", "Random Forest", "SVM", "XGBoost", "Logistic Reg."),
  Accuracy = c(89.36, 87.23, 85.11, 89.36, 82.98),
  Precision = c(0.87, 0.81, 0.80, 0.82, 0.75),
  Recall = c(0.81, 0.81, 0.75, 0.88, 0.75),
  F1 = c(0.84, 0.81, 0.77, 0.85, 0.75)
)

long_data <- results %>%
  pivot_longer(cols = c(Accuracy, Precision, Recall, F1),
               names_to = "Metric",
               values_to = "Value")
p_acc <- long_data %>% filter(Metric == "Accuracy") %>%
  ggplot(aes(x = Model, y = Value)) +
  geom_bar(stat = "identity", fill = "skyblue", width = 0.6, color = "#3A3A3A") +
  labs(title = "Accuracy", y = "Accuracy (%)", x = "") +
  coord_cartesian(ylim = c(80, 90)) +
  theme_minimal(base_size = 20) +
  theme(
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12),
    axis.title  = element_text(size = 14),
    plot.title  = element_text(size = 16, face = "bold")
  )

p_prec <- long_data %>% filter(Metric == "Precision") %>%
  ggplot(aes(x = Model, y = Value)) +
  geom_bar(stat = "identity", fill = "lightgreen", width = 0.6, color = "#3A3A3A") +
  labs(title = "Precision", y = "Precision", x = "") +
  coord_cartesian(ylim = c(0.7, 0.9)) +
  theme_minimal(base_size = 20) +
  theme(
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12),
    axis.title  = element_text(size = 14),
    plot.title  = element_text(size = 16, face = "bold")
  )
 

p_rec <- long_data %>% filter(Metric == "Recall") %>%
  ggplot(aes(x = Model, y = Value)) +
  geom_bar(stat = "identity", fill = "orange", width = 0.6, color = "#3A3A3A") +
  labs(title = "Recall", y = "Recall", x = "") +
  coord_cartesian(ylim = c(0.7, 0.9)) +
  theme_minimal(base_size = 20) +
  theme(
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12),
    axis.title  = element_text(size = 14),
    plot.title  = element_text(size = 16, face = "bold")
  )


p_f1 <- long_data %>% filter(Metric == "F1") %>%
  ggplot(aes(x = Model, y = Value)) +
  geom_bar(stat = "identity", fill = "plum", width = 0.6, color = "#3A3A3A") +
  labs(title = "F1-score", y = "F1-score", x = "") +
  coord_cartesian(ylim = c(0.7, 0.85)) +
  theme_minimal(base_size = 20) +
  theme(
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12),
    axis.title  = element_text(size = 14),
    plot.title  = element_text(size = 18, face = "bold")
  )


(p_acc | p_prec) /
(p_rec | p_f1)
```

\newpage

# Confusion Matrix Analysis

```{r cm-plot, include = FALSE}
plot_confusion_matrix <- function(pred, truth, title = "Confusion Matrix") {

  
  # desired order for the *labels* themselves
  desired_levels <- c("Nondemented", "Demented")
  
  # factors for prediction/truth in logical order for the x-axis
  truth_fac <- factor(truth, levels = desired_levels)
  pred_fac  <- factor(pred,  levels = desired_levels)
  
  # confusion table
  cm <- table(Predicted = pred_fac, True = truth_fac)
  cm_df <- as.data.frame(cm)
  
  # *** for the y-axis, put Nondemented on TOP, Demented on BOTTOM ***
  # (bottom -> top follows factor levels, so reverse them)
  cm_df$True <- factor(cm_df$True, levels = rev(desired_levels))
  
  neg_level <- desired_levels[1]  # "Nondemented"
  pos_level <- desired_levels[2]  # "Demented"
  
  cm_df <- cm_df %>%
    mutate(
      cell_type = case_when(
        True == "Nondemented" & Predicted == "Nondemented" ~ "TN",
        True == "Nondemented" & Predicted == "Demented"   ~ "FP",
        True == "Demented"    & Predicted == "Nondemented" ~ "FN",
        True == "Demented"    & Predicted == "Demented"    ~ "TP",
        TRUE ~ ""
      ),
      label = paste0(cell_type, " = ", Freq)
    )
  
  ggplot(cm_df, aes(x = Predicted, y = True, fill = Freq)) +
    geom_tile(color = "white") +
    geom_text(aes(label = label), size = 4, fontface = "bold") +
    scale_fill_gradient(low = "#FFFFCC", high = "#FF9933") +
    labs(
      title = title,
      x = "Predicted Label",
      y = "True Label"
    ) +
    theme_minimal(base_size = 14)
}

```

```{r tree-conf-matrix,include = FALSE, fig.cap = "Confusion matrix for decision tree."}
plot_confusion_matrix(tree_pred, y_test,
                      title = "Confusion Matrix – Decision Tree")

```


```{r rf-conf-matrix, include = FALSE, fig.cap = "Confusion matrix for random forest. "}
plot_confusion_matrix(rf_pred, y_test,
                      title = "Confusion Matrix – Random Forest")
```



```{r boost-conf-matrix, include = FALSE, fig.cap = "Confusion matrix for XGBoost. "}
plot_confusion_matrix(xgb_pred, y_test,
                      title = "Confusion Matrix – XGBoost.")
```


```{r SVM-conf-matrix, include = FALSE, fig.cap = "Confusion matrix for Support Vector Machine. "}
plot_confusion_matrix(svm_pred, y_test,
                      title = "Confusion Matrix – Support Vector Machine.")
```

```{r logit-conf-matrix, include = FALSE, fig.cap = "Confusion matrix for Logistic regression ", fig.pos='H'}
plot_confusion_matrix(logit_pred, y_test,
                      title = "Confusion Matrix – Logistic Regression")
```

The individual confusion matrices (Figure 2) provide the underlying count data that justifies the calculated performance metrics (Table 1). The confusion matrix data visually and numerically validates the conclusion that XGBoost is the optimal model. Its ability to minimize False Negatives (FN = 2) is the defining factor, ensuring the highest rate of true dementia case detection necessary for early intervention.


```{r combined-conf-matrices, echo = FALSE, message=FALSE, warning =FALSE, fig.height = 18, fig.width = 20, fig.pos = 't', fig.cap = "Individual Confusion Matrices for All Five Models."}

# library(your_plotting_package) # e.g., library(ggplot2) if your function uses it

# Assuming 'plot_confusion_matrix' generates a ggplot object, assign each plot to a variable.

# 1. Generate the individual plot objects and center their titles
center_title_plot <- function(p, title_text) {
  # Re-run your original plot function, then modify the title alignment
  p <- plot_confusion_matrix(p, y_test, title = title_text)
  
  # Check if p is a ggplot object before applying theme changes
  if (inherits(p, "ggplot")) {
    p <- p + theme(plot.title = element_text(hjust = 0.5), face = "bold")
  }
  return(p)
}

p1 <- center_title_plot(tree_pred, "Confusion Matrix – Decision Tree")
p2 <- center_title_plot(rf_pred, "Confusion Matrix – Random Forest")
p3 <- center_title_plot(xgb_pred, "Confusion Matrix – XGBoost")
p4 <- center_title_plot(svm_pred, "Confusion Matrix – Support Vector Machine")
p5 <- center_title_plot(logit_pred, "Confusion Matrix – Logistic Regression")

# 2. Use patchwork to combine them into a 2x3 layout
# The '+' operator adds plots; the 'plot_layout' sets the structure.
# 'plot_spacer()' is used to fill the empty 6th spot.
combined_plots <- (p1 + p2) / (p3 + p4) / (p5 + plot_spacer())

# print the combined plot
combined_plots
```

```{r, echo=FALSE}
compare_classifiers <- function(...) {
  # 1. Capture the models
  model_list <- list(...)
  
  # 2. Collect resampling results
  resamps <- caret::resamples(model_list)
  
  # 3. Print Statistical Comparison
  cat("--- Descriptive Statistics ---\n")
  print(summary(resamps))
  
  # 4. Statistical Significance Test
  cat("\n--- Hypothesis Testing (Model Differences) ---\n")
  diffs <- diff(resamps)
  print(summary(diffs))
  
  # 5. Visual Comparison
  # FIX: We must use the metrics that actually exist: ROC and Sens
  p <- bwplot(resamps, 
              metric = c("ROC", "Sens"), 
              main = "Model Comparison: MRI vs. No-MRI",
              scales = list(x = list(relation = "free")))
  
  print(p)
  
  return(resamps)
}


comparison_results_cross <- compare_classifiers(
  DT_Full = tree_model, 
  DT_NoMRI = tree_model1,
  RF_Full = rf_model,
  RF_NoMRI = rf_model1,
  XG_Full = xgb_model,
  XG_NoMRI = xgb_model1,
  SVM_Full = svm_model,
  SVM_NoMRI = svm_model1,
  Logit_Full = logit_model,
  Logit_NoMRI= logit_model1
  
)
summary(comparison_results_cross)
diffs_cross <- diff(comparison_results_cross)

median_roc_values <- summary(comparison_results_cross)$statistics$ROC[, "Median"]

print(median_roc_values)


```

```{r}
comparison_results <- compare_classifiers(
  DT_Full = tree_model2, 
  DT_NoMRI = tree_model3,
  RF_Full = rf_model2,
  RF_NoMRI = rf_model3,
  XG_Full = xgb_model2,
  XG_NoMRI = xgb_model3,
  SVM_Full = svm_model2,
  SVM_NoMRI = svm_model3,
  Logit_Full = logit_model2,
  Logit_NoMRI= logit_model3
  
)

summary(comparison_results)
diffs <- diff(comparison_results)
# Extract a summary table of all ROC (AUC) values
auc_summary <- summary(comparison_results)$statistics$ROC
# Print it to see Min, Max, Mean, and Median
```


